A Systematic Literature Review on Extensions to Role Based Access Control

### 2013-04-25 paper/sections/analysis.tex
fixing after spell check and grammar check based on word
spell check and grammar check based on word

\begin{itemize}
\setlength{\itemsep}{0.25pt}
\item The RBAC standard needs additional contextual information and constraints to develop fine-grained policies in practice.
\item The RBAC standard does not incorporate context. Therefore, RBAC belongs to static access control model, which may not capture changes in environments.
\item The RBAC standard does not support for various constraints such as temporal and spatial constraints to design sophisticated policies on demand.
\item The RBAC standard does not provide an abstraction to additional user-defined attributes	(e.g., task and team) and their association with existing attributes.
\item The RBAC standard has limitation on delegation and role hierarchy. For example, partial inheritance in role hierarchy needs to be developed.  
\item The RBAC standard needs to incorporate additional contextual information and constraints to develop fine-grained policies in practice.
\item The RBAC standard needs to incorporate additional attributes such as
task, team, purpose, organizational roles and collaborative activities over existing attributes. Moreover, new association between attributes
are introduced. 
\item RBAC model needs to improve existing features such delegation and role hierarchy to meet fine-grained access control. 
\end{itemize}

@@

\begin{itemize}
	\item Addition: an extended model incorporates additional elements with regards to context/constraints and their relations with existing elements of the RBAC standard. Models in this category does not change the core model. We found 23 papers in this category.
	\item Modification: an extended model incorporates modification of existing features in the RBAC standard. For example, changes of attribute relations such as role hierarchy is to show modification. For models in this category, changes occur within the core model. We found 1 papers in this category.
	\item Combination: an extended model is combination of RBAC and an another access control model, which uses model-specific attributes and their relations such as task, team, purpose, organizational roles. For this combination, new association between attributes of two models should be introduced. Models in this category requires a new association between attributes of different access control models to work together. We found 4 papers in this category.
\end{itemize}

@@

The RBAC standard presents four entities: roles, sessions, users, and permissions, and their relations. While this model is considered fundamental in any RBAC systems, this model can be extended to be applicable for emerging applications such as healthcare and mobile devices. 
In such environments, the RBAC standard needs additional constraints and dynamic context to support dynamic environments such as spatial and location. For this category, more than 80\% of papers are classified into ``Addition'' category. Besides, one paper~\cite{alam06:constraint} modifies relationship among elements to support partial inheritence. RBAC is shown to be combined with other access control models such as team-based access control and task-based access control~\cite{zhou2007team, oh2003task}.

@@

When designing and proposing a model targeted at a feature that is rooted in practical
usage by real software systems, bringing the model to life is strong evidence that the
proposed model can work in practice. The concept of authorization, and access control
is rooted in a business need. Thus, any access control model needs to be feasible
in the real world not just on paper. We analyzed the primary sources to see how many
proposed models actually had implementations associated with them.  And quantified the
type of implementation. Whether the implementation was for a real system, for a prototype
and/or used in a production environment.

@@

Table \ref{tab:implementations} shows the breakdown of implementations found within the primary sources.
Of the 28 papers surveyed, there was a lack of implementation with 20 of the paper providing no
mention of a implementation or prototype.  Of the remaining 8 papers that did mention an implementation, half 
were simply prototypes developed by the authors while the other half were claimed to be implemented within a real
system.

@@

The RBAC standard was designed with enterprises in mind such that when practitioners implemented RBAC into their systems
there would be a reasonable assurance being based off a well thought out model.  As extensions to the standard model
come along, thought and time should be given to how features and nuances of their models may impact implementation
in order to achieve the same goals as the original standard.  The primary sources should a significant lack of implementation
with over 70\% of the models having no notion of attempting to implement them.  The bare minimum, as 4 papers did, should be
a prototype implementation of the model for review by both practitioners and researchers. Of the models that produced an 
implementation within the enterprise world, two were from within the medical domain and two were implemented using web application technologies.

@@

RQ4: How are extension to RBAC evaluated theoretically and in practice?

@@

The 28 primary sources were examined for evidence that evaluations of the proposed model were presented by the model authors.
Further, we identified a set of evaluation types found within the each of the primary sources and provide below a list of the 
evaluation types and which primary sources provided which type.  In some cases a single primary source provided multiple evaluation
types.

@@

The primary source of validation a developer or practitioner may encounter is a qualitative discussion of real-world
scenarios and how the proposed model can tackle those situations. In rare cases, five primary sources, the model 
authors provide some discussion of how the RBAC standard model is deficient in tackling the scenario. In rarer cases, 
one paper from our results, a case study is performed to examine how the proposed model works in practice. Discussions
of how an extension model handles a real-world scenario provides developers and practitioners anecdotal evidence at best
for what types of situations the proposed model could handle. Further, by not providing a comparison to the RBAC standard
model, developers may be left implementing a more complex model to address their requirements when the standard model would
have sufficed.  Further, given the nature of access control models as grounded in application to enterprise implementations, 
case studies of a model in action provide developers with evidence that the model works as intended when applied.

@@

The predominant domain for which extension models have been generated for is that of the medical domain with 9 of 28 mentioning scenarios or requirements of that industry.
Mobile computing and enterprise workflows were each represented by 5 papers claiming to be influenced by the requirements for access control within these domains. The final set
of domains was pervasive computing environments and large-scale organizations with 3 each and web applications with 1.  There were 4 papers without any direct mention of a domain
since Aich et al. \cite{aich09:role} fall under both the medical domain and mobile computing.

@@

The medical domain produced the largest selection of papers when analyzing the domains influencing the proposals of extension models.  
Further, the authors noted that the categories associated with papers identifying the medical domain was not limited to one or two but cut across
each of the eight categories except for Organization. 
The cross-category nature of the medical domain papers appears indicative of the complex nature of medical applications and the requirements therein.
Given the growth of the research and development of medical applications over the past decade this result does not appear to be surprising. However,
the RBAC standard was originally created to reduce cost and increase interoperability - two goals of current regulation around the standardization
of electronic health record systems. The large number of proposed models, and the cross-category result stand in direct opposition of the goals
of both the RBAC standard and current regulations.

@@

\section{Generaliations} \label{sec:generalizations}

@@

Since NIST proposed RBAC standard using propositional logic, researchers describe extended RBAC models using propositional logic. Given context, propositional logic is used to evaluate true or false based on specified rules and access control logic. As access control is typically evaluated to either true (Permit) or false (Deny) based on predicates, propositional logic is sufficient to describe key ideas and definitions. We found that OO papers of RBAC extensions use propositional logic to describe its extended model.
However, propositional logic has limitations. While this logic is simple, this logic does not support for reasoning about RBAC, which helps reduce the administrative complexity of associations such as user- role associations. An alternative logic to describe RBAC is first-order logic. This logic is sufficient to describe RBAC and extended RBAC. Moreover, this logic supports for concise and elegant formulation of the RBAC model and its relation.  First-Order logic is expressive enough to concisely represent access control systems. First-Order logic uses relations, variables, and quantifiers.

>>>>

\begin{itemize}
\setlength{\itemsep}{0.25pt}
\item The RBAC standard needs additional contextual information and constraints to develop fine-grained policies in practice.
\item The RBAC standard does not incorporate context. Therefore, RBAC belongs to static access control model, which may not capture changes in environments.
\item The RBAC standard does not support for various constraints such as temporal and spatial constraints to design sophisticated policies on demand.
\item The RBAC standard does not provide an abstraction to additional user-defined attributes	(e.g., task and team) and their association with existing attributes.
\item The RBAC standard has limitation on delegation and role hierarchy. For example, partial inheritance in role hierarchy needs to be developed.  
\item The RBAC standard needs to incorporate additional contextual information and constraints to develop fine-grained policies in practice.
\item The RBAC standard needs to incorporate additional attributes such as
task, team, purpose, organizational roles and collaborative activities over existing attributes. Moreover, new association between attributes
is introduced. 
\item RBAC model needs to improve existing features such delegation and role hierarchy to meet fine-grained access control. 
\end{itemize}

@@

\begin{itemize}
	\item Addition: The extension model incorporates additional elements with regards to context/constraints and their relations with existing elements of the RBAC standard. Models in this category do not change the RBAC standard model. We found 23 papers in this category.
	\item Modification: The extension model incorporates modification of existing features in the RBAC standard. For example, changes of attribute relations such as role hierarchy are to show modification. For models in this category, changes occur within the the RBAC standard model. We found one paper in this category.
	\item Combination: The extension model is combination of RBAC and another access control model, which uses model-specific attributes and their relations such as task, team, purpose, organizational roles. For this combination, new association between attributes of two models should be introduced. Models in this category require a new association between attributes of different access control models to work together. We found four papers in this category.
\end{itemize}

@@

The RBAC standard presents four entities: roles, sessions, users, and permissions, and their relations. While this model is considered fundamental in any RBAC systems, this model can be extended to be applicable for emerging applications such as healthcare and mobile devices. 
In such environments, the RBAC standard needs additional constraints and dynamic context to support dynamic environments such as spatial and location. For this category, more than 80\% of papers are classified into ``Addition'' category. Besides, one paper~\cite{alam06:constraint} describes modification of existing relationship among elements to support partial inheritance. The RBAC standard is shown to be combined with other access control models such as team-based access control and task-based access control~\cite{zhou2007team, oh2003task}.

@@

When designing and proposing a model targeted at a feature that is rooted in practical
usage by real software systems, bringing the model to life is strong evidence that the
proposed model can work in practice. The concept of authorization and access control
is rooted in a business need. Thus, any access control model needs to be feasible
in the real world not just on paper. We analyzed the primary sources to see how many
proposed models actually had implementations associated with them.  Then, we quantified
types of implementations. Whether the implementation was for a real system, for a prototype
and/or used in a production environment.

@@

Table \ref{tab:implementations} shows the breakdown of implementations found within the primary sources.
Of the 28 papers surveyed, there was a lack of implementation with 20 of the paper providing no
mention of an implementation or prototype.  Of the remaining eight papers that did mention an implementation, half 
were simply prototypes developed by the authors while the other half were claimed to be implemented within a real
system.

@@

The RBAC standard was designed with enterprises in mind such that when practitioners implemented RBAC into their systems
there would be a reasonable assurance being based off a well thought out model.  As extensions to the standard model
come along, thought and time should be given to how features and nuances of their models may impact implementation
in order to achieve the same goals as the original standard.  The primary sources should a significant lack of implementation
with over 70\% of the models having no notion of attempting to implement them.  The bare minimum, as four papers did, should be
a prototype implementation of the model for review by both practitioners and researchers. Of the models that produced an 
implementation within the enterprise world, two were from within the medical domain and two were implemented using web application technologies.

@@

RQ4: How are extensions to RBAC evaluated theoretically and in practice?

@@

The 28 primary sources were examined for evidence that evaluations of the proposed model were presented by the model authors.
Further, we identified a set of evaluation types found within  each of the primary sources and provide below a list of the 
evaluation types and which primary sources provided which type.  In some cases a single primary source provided multiple evaluation
types.

@@

The primary source of validation a developer or practitioner may encounter is a qualitative discussion of real-world
scenarios and how the proposed model can tackle those situations. In rare cases, five primary sources, the model 
authors provide some discussion of how the RBAC standard model is deficient in tackling the scenario. In rare cases, 
one paper from our results, a case study is performed to examine how the proposed model works in practice. Discussions
of how an extension model handles a real-world scenario provides developers and practitioners anecdotal evidence at best
for what types of situations the proposed model could handle. Further, by not providing a comparison to the RBAC standard
model, developers may be left implementing a more complex model to address their requirements when the standard model would
have sufficed.  Further, given the nature of access control models as grounded in application to enterprise implementations, 
case studies of a model in action provide developers with evidence that the model works as intended when applied.

@@

The predominant domain for which extension models have been generated for is that of the medical domain with nine of 28 mentioning scenarios or requirements of that industry.
Mobile computing and enterprise workflows were each represented by five papers claiming to be influenced by the requirements for access control within these domains. The final set
of domains was pervasive computing environments and large-scale organizations with three each and web applications with one.  There were four papers without any direct mention of a domain
since Aich et al. \cite{aich09:role} fall under both the medical domain and mobile computing.

@@

The medical domain produced the largest selection of papers when analyzing the domains influencing the proposals of extension models.  
Further, the authors noted that the categories associated with papers identifying the medical domain were not limited to one or two but cut across
each of the eight categories except for Organization. 
The cross-category nature of the medical domain papers appears indicative of the complex nature of medical applications and the requirements therein.
Given the growth of the research and development of medical applications over the past decade this result does not appear to be surprising. However,
the RBAC standard was originally created to reduce cost and increase interoperability - two goals of current regulation around the standardization
of electronic health record systems. The large number of proposed models, and the cross-category result stand in direct opposition of the goals
of both the RBAC standard and current regulations.

@@

\section{Generalizations} \label{sec:generalizations}

@@

Since NIST proposed RBAC standard using propositional logic, researchers describe extended RBAC models using propositional logic. Given context, propositional logic is used to evaluate true or false based on specified rules and access control logic. As access control is typically evaluated to either true (Permit) or false (Deny) based on predicates, propositional logic is sufficient to describe key ideas and definitions. We found that 28 papers of RBAC extensions use propositional logic to describe its extended model.
However, propositional logic has limitations. While this logic is simple, this logic does not support for reasoning about RBAC, which helps reduce the administrative complexity of associations such as user- role associations. An alternative logic to describe RBAC is first-order logic. This logic is sufficient to describe RBAC and extended RBAC. Moreover, this logic supports for concise and elegant formulation of the RBAC model and its relation.  First-Order logic is expressive enough to concisely represent access control systems. First-Order logic uses relations, variables, and quantifiers.

### 2013-04-25 paper/sections/categorization.tex
fixing after spell check and grammar check based on word
spell check and grammar check based on word

We offer definitions for each of the 8 observed categories based on the data extracted from the primary sources and the English definitions for each noun-phrase. 
Some category descriptors contain abbreviations in parenthesis that match the shortened name found within Table \ref{tab:categorization}.

@@

  \item \textbf{Task}: The extension model provides a task entity which are associated with permissions and roles. A task is a fundamental unit of a business activity. Different from core RBAC, in task-role-based access control model, roles are not directly associated with permissions. Roles are first associated with tasks, which are associated permissions. For example, the employee role is associated with a task, which is to write a report. Then, this task is associated with a permission.

@@

The 28 primary sources produced a set of 8 heirarchical categories. Table \ref{tab:categorization} summarizes each primary source under their designated
category and further, displays the perceived heirarchy of the categories. 
The Spatial and Temporal categories were treated as subsets of the broader category of Spatio-Temporal since this category encompasses them individually and
the Spatio-Temporal category contained more primary sources that the Spatial or Temporal categories alone.

>>>>

We offer definitions for each of the eight observed categories based on the data extracted from the primary sources and the English definitions for each noun-phrase. 
Some category descriptors contain abbreviations in parenthesis that match the shortened name found within Table \ref{tab:categorization}.

@@

  \item \textbf{Task}: The extension model provides task elements which are associated with permissions and roles. A task is a fundamental unit of a business activity. Different from core RBAC, in task-role-based access control model, roles are not directly associated with permissions. Roles are first associated with tasks, which are associated permissions. For example, the employee role is associated with a task, which is to write a report. Then, this task is associated with a permission element.

@@

The 28 primary sources produced a set of eight hierarchical categories. Table \ref{tab:categorization} summarizes each primary source under their designated
category and further, displays the perceived hierarchy of the categories. 
The Spatial and Temporal categories were treated as subsets of the broader category of Spatio-Temporal since this category encompasses them individually and
the Spatio-Temporal category contained more primary sources that the Spatial or Temporal categories alone.

### 2013-04-25 paper/sections/discussion.tex
fixing after spell check and grammar check based on word
spell check and grammar check based on word

A small percentage of primary sources presented substantial evaluations or implementations of their model in practice.  This in turn leaves researchers and practioners little data from which to compare one model to the next. And given the wide swath of popluar and critical domains being targeted by RBAC extension models, we propose a set of metrics by which RBAC extension models should be evaluated. Hu, Vincent C. and Scarfone, Karen provide metrics for access control system evaluation or a security system in general~\cite{hu2012:NIST7874}. Hu, Vincent C. and Scarfone, Karen collect metrics, which can be applicable for not only access control models, but also requirements, system implementation, and extended application such as verification and testing. From these metrics, and guided by the data surrounding how extension models are currently evaluated, we select metrics that can be applicable for evaluation of extended RBAC models as follows:

\begin{itemize}
	\item \textbf{Ease of Privilege Assignments}: measures number of steps for assigning, changing, or deleting a privilege for users and roles when a user or administrator manage RBAC. If a model requires more steps for such a given task, there is a potential increase in human or system error due to increased complexity. While extended RBAC may have similar features to that of the RBAC standard model, we measure worst cases after extended, and the gap between an extension model and the RBAC standard model.

@@

We recommend these metrics be incorporated in any extension models presented hence forth. A future work could include providing metric measurements for the primary sources prsented here. 
A developer could attemot to apply these metrics to an extension model not already evaluated when attempting to choose a model. 
These metrics focus on comparing specific features support of access control models.

>>>>

A small percentage of primary sources presented substantial evaluations or implementations of their model in practice.  This in turn leaves researchers and practitioners little data from which to compare one model to the next. And given the wide swath of popular and critical domains being targeted by RBAC extension models, we propose a set of metrics by which RBAC extension models should be evaluated. Hu and Scarfone provide metrics for access control system evaluation or a security system in general~\cite{hu2012:NIST7874}. They collect metrics, which can be applicable for not only access control models, but also requirements, system implementation, and extended application such as verification and testing. From these metrics, and guided by the data surrounding how extension models are currently evaluated, we select metrics that can be applicable for evaluation of extended RBAC models as follows:

\begin{itemize}
	\item \textbf{Ease of Privilege Assignments}: measures number of steps for assigning, changing, or deleting a privilege for users and roles when a user or administrator manages RBAC. If a model requires more steps for such a given task, there is a potential increase in human or system error due to increased complexity. While extended RBAC may have similar features to that of the RBAC standard model, we measure worst cases after extended, and the gap between an extension model and the RBAC standard model.

@@

We recommend these metrics be incorporated in any extension models presented hence forth. A future work could include providing metric measurements for the primary sources presented here. 
A developer could attempt to apply these metrics to an extension model not already evaluated when attempting to choose a model. 
These metrics focus on comparing specific features support of access control models.

### 2013-04-25 paper/sections/process.tex
Minor updates to process and categories.
Committing updates based on comments from reading group.
Adding greybox. Adding results subsections for each RQ. Fixing last table captions to be on top. Edits to categories.
fixing after spell check and grammar check based on word
spell check and grammar check based on word

\section{Process} \label{sec:process}

@@

For the first phase of our systematic literature review, the authors developed a search strategy for finding papers.  The search strategy was executed by an automated comprehensive search taking as input a set of academic search engines and a list of search terms.  The search was performed by applying each search term to each engine incrementally until a stopping criteria was met. Table \ref{tab:search_results} lists the four search engines along the left most column with the three search term across the top row along with the papers for each criteria and engine combination.  The search algorithm was performed as follows:

\begin{enumerate}
\setlength{\itemsep}{0.25pt}
\item Call to search engine with current search position and current search term.
\item Parse results and extract paper title, authors and year of publication.
\item Compare results against stopping criteria:
	\begin{itemize}
	\item If the size of the result set is greater than or equal to 1000 then stop.
	\item If the last ten results did not contain the search term phrase within the tite then stop.
	\end{itemize}
\item If stopping criteria not met, increment search position and go back to step one.
\end{enumerate}

@@

\begin{table}
\centering
\begin{tabular}{|p{3.5cm}|p{1.25cm}|p{4.25cm}|p{4.25cm}|p{1cm}|}

@@

\end{tabular}
\caption{Paper counts after applying search strategy}
\label{tab:search_results}
\end{table}

@@

\begin{itemize}
\setlength{\itemsep}{0.25pt}
\item Does the abstract mention a proposed model?
\item Does the abstract mention extension of role-based access control?
\item Does the abstract mention either an implementation, evaluation or domain for their model?
\end{itemize}

@@

Question 1 was a definitive exclusion criteria as any paper that failed in the affirmative was rejected.  Questions 3 and 4 were marked as conditional includes given that they were connected in making a decision.  A paper that met Question 3 but not 4, or met 4 but not 3 would be included since in some cases the real world examples served as research evaluations and without this conditional include the paper list size would be too small to be significant.

\begin{table}
\centering
\begin{tabular}{ c c | c | c | c | }
\cline{3-5}

@@

\multicolumn{1}{ c| }{} & Overlap & 249 & 51 & 24 \\ \cline{2-5}
\multicolumn{1}{ c| }{} & Disjoint & 232 & 137 & 64 \\ \cline{2-5}
\multicolumn{1}{ c| }{} & Rejected & 208 & 116 & 59 \\ \cline{2-5}
\multicolumn{1}{ c| }{} & Retained & 41 & 21 & 5 \\ \cline{2-5}
\multicolumn{1}{ c| }{} & \textbf{Total} & 290 & 72 & 29 \\ \cline{2-5}

\end{tabular}
\caption{Elimination Rounds}
\label{tab:search_results}
\end{table}

>>>>

\section{Methodology and Process} \label{sec:process}

@@

For the first phase of our systematic literature review, the authors developed a search strategy for finding papers.  The search strategy was executed by an automated comprehensive search taking as input a set of academic search engines and a list of search terms.  The search was performed by applying each search term to each engine incrementally until stopping criteria were met. Table \ref{tab:search_results} lists the four search engines along the left most column with the three search term across the top row along with the papers for each criteria and engine combination.  The search algorithm was performed as follows:

\begin{enumerate}
\setlength{\itemsep}{0.25pt}
\item Call to search engine with current search position and current search term.
\item Parse results and extract paper title, authors and year of publication.
\item Compare results against stopping criteria:
	\begin{itemize}
	\item If the size of the result set is greater than or equal to 1000 then stop.
	\item If the last ten results did not contain the search term phrase within the title then stop.
	\end{itemize}
\item If stopping criteria not met, increment search position and go back to step one.
\end{enumerate}

@@

\begin{table}
\centering
\caption{Paper counts after applying search strategy}
\begin{tabular}{|p{3.5cm}|p{1.25cm}|p{4.25cm}|p{4.25cm}|p{1cm}|}

@@

\end{tabular}
\label{tab:search_results}
\end{table}

@@

\begin{itemize}
\setlength{\itemsep}{0.25pt}
\item Does the abstract mention a proposed model?
\item Does the abstract mention extension of role-based access control?
\item Does the abstract mention an implementation, evaluation or domain for their model?
\end{itemize}

@@

Question 1 was a definitive exclusion criterion as any paper that failed in the affirmative was rejected.  Questions 3 and 4 were marked as conditional includes given that they were connected in making a decision.  A paper that met Question 3 but not 4, or met 4 but not 3 would be included since in some cases the real world examples served as research evaluations and without this conditional include the paper list size would be too small to be significant.

\begin{table}
\centering
\caption{Elimination Rounds}
\begin{tabular}{ c c | c | c | c | }
\cline{3-5}

@@

\multicolumn{1}{ c| }{} & Overlap & 249 & 51 & 24 \\ \cline{2-5}
\multicolumn{1}{ c| }{} & Disjoint & 232 & 137 & 64 \\ \cline{2-5}
\multicolumn{1}{ c| }{} & Rejected & 208 & 116 & 59 \\ \cline{2-5}
\multicolumn{1}{ c| }{} & Retained & 41 & 21 & 5 \\ \cline{2-5}
\multicolumn{1}{ c| }{} & \textbf{Total} & 290 & 72 & 28 \\ \cline{2-5}

\end{tabular}
\label{tab:search_results}
\end{table}
